### 2021-05-09 papers/spherical-caps/revised/response-to-referees.tex

\bigskip

\centerline{\textbf{Response to Referee 1}}

\begin{itemize}[parsep=1em,leftmargin=1em]

\item\comment{If there is a key technical issue that the authors overcome to allow this construction to proceed it is not obvious from the paper, meaning that the reader is left with the impression that, of course, the previously developed sparse spectral methods on 2D planar domains should extend}

Handling partial differential operators on the sphere is significantly more challenging than partial differential operators in Euclidean space. This is similar to the significantly increased challenge faced in solving PDEs on spheres (e.g.  [22]) compared with disks (e.g. [21]). Note further that the construction for 2D planar domains did not take advantage of rotational symmetry.

@@

\item \comment{The authors spend an inadaquate amount of time discussing other approaches for solving this problem and comparing their methodology (e.g., closest point method, radial basis functions). What is the complexity convergence behavior of other approaches that can handle spherical caps?}

We have added  a brief discussion of the closest point method. Note that the closest point method does not achieve spectral accuracy so is not really comparable to our approach. Further it does not take advantage of rotational symmetry which means that direct solvers will not have optimal complexity.   We have chosen to omit discussion on radial basis functions: while they have been used with great success on a full sphere, we are not aware of references that consider their usage on spherical caps. If the referee can point us to relevant references that would be very much  appreciated.

@@

\item \comment{Definition 2: It should be mentioned somewhere that OPs should defined by the condition that they are orthogonal to all lower degree polynomials.}

Added a comment before Definition 2 to explain this, as well as an extra line in Definition 2. Please note that in more than 1D this condition on its  own is only sufficient to define the span of the degree $n$ polynomials: the choice of basis for the degree $n$ polynomials is not unique. Definition 2 therefore determines our choice of basis.

@@

\centerline{\textbf{Response to Referee 2}}

\begin{itemize}[parsep=1em,leftmargin=1em]

\item \comment{One shortcoming of this paper is that it has no discussion of the condition number of the linear systems to be solved. I suspect that the condition numbers can get large, but if so, can the matrices be preconditioned in a similar way the Olver-Townsend ultraspherical spectral method (so that it has bounded condition number)?}

Added small subsection and plots to show the condition numbers of each m-block of Laplacian operator, along with a preconditioned Laplacian (where the preconditioner $P$ is the inverse of the diagonal of the Laplacian). While the Laplacian itself has a growing condition number, the trivial preconditioning makes it well-conditioned.

@@

\item \comment{If the condition numbers do not get large, can you prove it, indicate why, or give numerical evidence?}

We have added numerical evidence demonstrating that the discretisation is well-conditioned under a simple preconditioner. To prove this rigorously is likely to be quite involved as we believe it would involve determining the asymptotics of the operators. Even in the case of the full disk [21], where the entries of operators are known explicitly, full condition number analysis is still open, as one needs to handle singularly perturbed operators for the large $m$ modes.

@@

\item \comment{The paper only includes numerical experiments involving zero Dirichlet boundary conditions. Please could the authors clarify in the manuscript if there are any roadblocks to boundary conditions beyond the zero Dirichlet ones demonstrated in Section 5?}

Added comment a small subsection to Section 5 to address this. Neumann conditions should be doable by writing the equation as a system of PDEs, as shown in [13] on the triangle.

@@

\item \comment{Is the code used to produce the numerical experiments going to be made publicly available? For example in the ApproxFun Julia package?}

The package that makes up the framework for expanding functions, obtaining the operator matrices etc. is purely experimental at this stage, but is publicly available, and a link has been added to the paper.

### 2021-05-20 papers/thesis/full/ch1-introduction/introduction.tex
Further edits to ch1 and 2. Added images for examples and spy plots of jacobi mats.

Spherical Harmonics are of course a long-established and famous group of functions defined on the surface of a sphere with certain useful properties (for example, they are orthogonal to each other on the unit sphere and form a basis for expanding functions defined on the sphere) and as a result are widely used in many fields including computer graphics (e.g. \cite{moon2008efficient, sloan2013efficient}), astrophysics (e.g. \cite{vasil2019tensor}), quantum theory \cite{varshalovich1988quantum} and geosciences \bstodo{cite applications of SHs}. Notably, they are also used as basis functions for the spectral transform method that makes up part of the model in the Integrated Forecasting System (IFS), which is used by ECMWF for their forecasts \cite{wedi2013fast}. While the whole sphere spectral method using the Spherical Harmonics has been successful for numerous years \cite{williamson2007evolution}, there is a drawback in the parallel scalability bottleneck that arises from the global spectral transform, which is expected to inhibit future performance of the IFS \cite{ecmwf2020scalability, wedi2013fast, diamantakisecmwf}.

@@

There are also potential applications in physics too, particularly in astrophysics, where solving equations on the sphere and working in spherical geometries is also desirable \cite{vasil2019tensor}\bstodo{cite more, see p2 of Vasil and the references}. Recently, a method for computing tensor fields in spherical coordinates using Jacobi polynomials has been proposed \cite{vasil2019tensor}. In this work the authors present the method for both the surface of the unit sphere and the three-dimensional generalisation of the unit ball. 
\bstodoinline{Describe/summarise the Vasil 2019 Tensor paper approach. Compare to our approach (spectral method with OPs on spherical sub-domains).}

@@

>>>>

Spherical Harmonics are of course a long-established and famous group of functions defined on the surface of a sphere with certain useful properties (for example, they are orthogonal to each other on the unit sphere and form a basis for expanding functions defined on the sphere) and as a result are widely used in many fields including computer graphics (e.g. \cite{moon2008efficient, sloan2013efficient}), astrophysics (e.g. \cite{vasil2019tensor}), quantum theory (e.g. \cite{varshalovich1988quantum}), geosciences (e.g. \cite{hollerbach2013parity}) and meteorology (e.g. \cite{evans1998spherical, rubinstein2015scalar}) \bstodo{cite applications of SHs}. Notably, they are also used as basis functions for the spectral transform method that makes up part of the model in the Integrated Forecasting System (IFS), which is used by ECMWF for their forecasts \cite{wedi2013fast}. While the whole sphere spectral method using the Spherical Harmonics has been successful for numerous years \cite{williamson2007evolution}, there is a drawback in the parallel scalability bottleneck that arises from the global spectral transform, which is expected to inhibit future performance of the IFS \cite{ecmwf2020scalability, wedi2013fast, diamantakisecmwf}.

@@

There are also potential applications in physics too where solving equations on the sphere and working in spherical geometries is also desirable (e.g. \cite{beyer2014numerical, varshalovich1988quantum, slevinsky2018spectral})\bstodo{cite more, see p2 of Vasil and the references}, particularly in astrophysics \cite{vasil2019tensor, reinecke2013libsharp}.

Recently, a method for computing tensor fields in spherical coordinates using Jacobi polynomials has been proposed \cite{vasil2019tensor}. In this work the authors present the method for both the surface of the unit sphere and the three-dimensional generalisation of the unit ball. Their method involves using a spectral basis to represent functions too, choosing for the angular part of the basis to be the spin-weighted spherical harmonics (of which the spherical harmonics are a special case, with spin $0$) in spherical coordinates. By using these as their basis functions, they are able to derive sparse relations for the Laplacian operator, as well as for multiplication of the basis by $\cosphi$ and $\sinphi$ (where $\varphi$ is the polar angle from the $z$ axis in cartesian coordinates) which lead to operators for operations with angular dependencies involving these.

On the other hand, we aim to propose a strictly orthogonal \textit{polynomial} basis, and derive sparse operators for multiplication by the cartesian coordinate axes, which can in turn lead to operators for multiplication by trigonometric functions too. By utilising orthogonal polynomials, we can develop a sparse spectral method on subdomains of the sphere surface (e.g. the spherical cap) using our knowledge of other geometries and methods involving OPs developed for them.

@@

Finally, we summarise the work we have presented and detail avenues for future directions that one could take it in Chapter 5.
